Gravitational field calculation in spherical coordinates using variable densities in depth

### 2018-11-27 manuscript/manuscript.tex
Add missing citation of Uieda2016

\begin{summary}
We present a new methodology to compute the gravitational fields generated by
tesseroids (spherical prisms) whose density varies with depth according to
an arbitrary continuous function.
It approximates the gravitational fields through the Gauss-Legendre Quadrature along
with two discretization algorithms that automatically control its accuracy by adaptively
dividing the tesseroid into smaller ones.
The first one is a preexisting adaptive discretization algorithm that reduces
the errors due to the distance between the tesseroid and the computation
point.
The second is a new density-based discretization algorithm that
decreases the errors introduced by the variation of the density function with depth.
The amount of divisions made by each algorithm is indirectly controlled
by two parameters: the distance-size ratio and the delta ratio.
We have obtained analytical solutions for a spherical shell with radially variable
density and compared them to the results of the numerical model for linear and
exponential density functions.
These comparisons allowed us to obtain optimum values for the distance-size and
delta ratios that yield an accuracy of 0.1\% of the analytical solutions.
The resulting optimal values of distance-size ratio for the gravitational potential, its
gradient, and Marussi tensor are 1, 2 and 8, respectively.
A delta ratio of 0.2 is needed for the computation of the gravitational potential and
its gradient components, while a value of 0.01 must be used for the Marussi tensor
components.
Lastly, we apply this new methodology to model the Neuqu\'en Basin, a foreland basin in
Argentina with a maximum depth of over 5000 m, using an exponential density function.
\end{summary}

@@

The main challenge of the GLQ integration is loss of accuracy when the computation point
gets closer to the tesseroid \citep{Ku1977}.
\citet{Uieda2016} built on the adaptive discretization algorithm of \citet{Li2011}
to automatically obtain integration results with 0.1\% accuracy.
The algorithm consist in recursively splitting the tesseroid into smaller ones when a
threshold is exceeded,
namely when the normalized distance to the computation point is greater than a
``distance-size ratio'' ($D$).
\citet{Uieda2016} have also obtained standard values of $D$
for the gravitational potential, acceleration, and gradient tensor components
by comparing the numerical model with the fields generated by a spherical shell.

Two recent papers present other approaches for homogeneous tesseroids gravitational
fields computation and incorporate methodologies for variable density tesseroid with
specific density profiles.
\citet{Fukushima2018} analytically integrates the volumetric integral for the
gravitational potential in the radial direction, obtaining a surface integral, which is
then numerically solved by applying a conditionally splitting and the double exponential
quadrature rule. The gradient of the potential and the Marussi tensor components are
then computed by finite differences. \citet{Fukushima2018} also generalizes his method
to an arbitrary polynomial density profile on the radial coordinate.
\citet{Lin2018} made a comparison of the different methodologies for homogeneous
tesseroid gravitational fields computation by applying different discretizations
algorithms to each one of them and developed a combined method.
For computation points near the tesseroid they apply a GLQ numerical integration with
a two dimensional adaptive discretization algorithm, i.e. only in the horizontal
dimension. But, if the computation point is farther than a certain truncation distance,
then a second order Taylor series approximation is applied along with the regular
subdivision developed by \citet{Grombein2013}.
They also introduce a variation of their combined method to compute the gravitational
fields generated by tesseroids with linear density on the radial coordinate.

In summary, the bibliography is offering different methodologies for computing
gravitational fields of homogeneous tesseroids or with certain specific density
functions in depth.
Although any density profile could be approximated by one of these specific density
functions, one should acknowledge the error that such approximation would propagate into
the gravitational field in order to guarantee certain degree of accuracy.

We present a new algorithm for computing the gravitational fields generated by any
tesseroid with any arbitrary continuous density function on any external point.
It is based on the GLQ approximation and the adaptive discretization of
\citet{Uieda2016}, which we extend to include a new density-based discretization step.
In order to ensure the accuracy of the numerical approximation we have
determined optimal values for the controlling parameters by
comparing the numerical approximation with analytical solutions for
spherical shells.
Finally, we applied the methodology to model the Neuqu\'en basin, Argentina, using
tesseroids with an exponentially increasing density with depth.

@@

We define an external computation point $P(r, \phi, \lambda)$ in a geocentric spherical
coordinate system at a radius $r$, geocentric latitude $\phi$, and longitude $\lambda$
where the gravitational fields are going to be calculated.
The first and second derivatives of the gravitational potential are taken with
respect to the local north-oriented Cartesian coordinate system of $P$
(Fig.~\ref{fig:tesseroid}).
\citet{Grombein2013} provide efficient formulations for the volume integrals of the
gravitational potential and its first and second derivatives of a tesseroid with
homogeneous density.
Here, we will assume that the tesseroid has a density varying with $r$ according to an
arbitrary function $\rho(r)$.
Thus, the integrals for the gravitational fields are slightly modified to

@@

\noindent $(r_i, \phi_j, \lambda_k)$ are the quadrature nodes, and $W_i^r$ are the
quadrature weights.

@@

On the first iteration, the ratio $L_r/L_r^\text{orig} = 1$ because the tesseroid being
divided is the original one.
For future iterations, the ratio will be progressively smaller than one as the
tesseroids get smaller.
This is intended to limit the number of divisions to the ones that will
significantly reduce the numerical error:
dividing a large tesseroid with a small $\text{max}\{ \Delta \rho(r') \}$ would
improve the integration accuracy more than dividing a small tesseroid with a
higher $\text{max}\{ \Delta \rho(r') \}$.
We also do not divide tesseroids with $L_r < 1$ mm because any integration errors would
be negligible due to the small mass of the tesseroid..

@@

\noindent
Because of the rotational symmetry along the $z$ axe, the integration in $\lambda'$ is
straightforward:

@@

\begin{equation}
    g_{xx}(r) = g_{yy}(r) = -\frac{V_\text{sh}(r)}{r^2}
\end{equation}

@@

From Eq. \ref{eq:shell-pot} we can obtain expressions for gravitational potential for
different density functions.
For a linear density function

>>>>

\begin{summary}
We present a new methodology to compute the gravitational fields generated by
tesseroids (spherical prisms) whose density varies continuously with depth according to
an arbitrary function.
It approximates the gravitational fields through the Gauss-Legendre Quadrature along
with two discretization algorithms that automatically control its accuracy by adaptively
dividing the tesseroid into smaller ones.
The first one is a preexisting adaptive discretization algorithm that reduces
the errors due to the distance between the tesseroid and the computation
point.
The second is a new density-based discretization algorithm that
decreases the errors introduced by the variation of the density function with depth.
The amount of divisions made by each algorithm is indirectly controlled
by two parameters: the distance-size ratio and the delta ratio.
We have obtained analytical solutions for a spherical shell with radially variable
density and compared them to the results of the numerical model for linear and
exponential density functions.
These comparisons allowed us to obtain optimum values for the distance-size and
delta ratios that yield an accuracy of 0.1\% of the analytical solutions.
The resulting optimal values of distance-size ratio for the gravitational potential, its
gradient, and Marussi tensor are 1, 2 and 8, respectively.
A delta ratio of 0.2 is needed for the computation of the gravitational potential and
its gradient components, while a value of 0.01 must be used for the Marussi tensor
components.
Lastly, we apply this new methodology to model the Neuqu\'en Basin, a foreland basin in
Argentina with a maximum depth of over 5000 m, using an exponential density function.
\end{summary}

@@

The main challenge of the GLQ integration is loss of accuracy when the computation point
gets closer to the tesseroid \citep{Ku1977}.
\citet{Uieda2016} built on the adaptive discretization algorithm of \citet{Li2011}
to automatically obtain integration results with 0.1\% accuracy.
The algorithm consist in recursively splitting the tesseroid into smaller ones when a
threshold is exceeded,
namely when the normalized distance to the computation point is greater than a
``distance-size ratio'' ($D$).
\citet{Uieda2016} have also obtained standard values of $D$
for the gravitational potential, acceleration, and gradient tensor components
by comparing the numerical model with the analytical solution of the fields generated by
a spherical shell.

We present a new algorithm for computing the gravitational fields generated by any
tesseroid with an arbitrary continuous density function on any external point.
It is based on the GLQ approximation and the adaptive discretization of
\citet{Uieda2016}, which we extend to include a new density-based discretization step.
In order to ensure the accuracy of the numerical approximation we have
determined optimal values for the controlling parameters by
comparing the numerical approximation with analytical solutions for
spherical shells.
Finally, we applied the methodology to model the Neuqu\'en basin, Argentina, using
tesseroids with an exponentially increasing density with depth.

@@

Consider an arbitrary tesseroid in a geocentric spherical coordinate system defined by
pairs of geocentric latitudinal ($\phi_1$, $\phi_2$), longitudinal ($\lambda_1$,
$\lambda_2$) and radial ($r_1$, $r_2$) boundaries.
We define an external computation point $P(r, \phi, \lambda)$ located at radius $r$,
geocentric latitude $\phi$, and longitude $\lambda$ where the gravitational fields the
tesseroid generates are going to be calculated.
The first- and second-order derivatives of the gravitational potential are taken with
respect to the local north-oriented Cartesian coordinate system of $P$
(Fig.~\ref{fig:tesseroid}).
\citet{Grombein2013} provide efficient formulations for the volume integrals of the
gravitational potential and its first- and second-order derivatives of a tesseroid with
homogeneous density.
Here, we will assume that the tesseroid has a density varying with $r$ according to an
arbitrary function $\rho(r)$.
Thus, the integrals for the gravitational fields are slightly modified to

@@

\noindent $f(r', \phi', \lambda')$ is the non-density dependent portion of the
integration kernels, $(r_i, \phi_j, \lambda_k)$ are the quadrature nodes, and
$W_i^r$, $W_j^\phi$, $W_k^\lambda$ are the quadrature weights.
The GLQ nodes and weights can be calculated as \citet[p.~391]{Hildebrand1987} (see
\citet{Uieda2016} for nodes and weights definitions on similar notation as this paper).

@@

On the first iteration, the ratio $L_r/L_r^\text{orig} = 1$ because the tesseroid being
divided is the original one.
For further iterations, the ratio will be progressively smaller than one as the
tesseroids get smaller.
This is intended to limit the number of divisions to the ones that will
significantly reduce the numerical error:
dividing a large tesseroid with a small $\text{max}\{ \Delta \rho(r') \}$ would
improve the integration accuracy more than dividing a small tesseroid with a
higher $\text{max}\{ \Delta \rho(r') \}$.
We also do not divide tesseroids with $L_r < 1$ mm because any integration errors would
be negligible due to the small mass of the tesseroid.

@@

\noindent
Because of the rotational symmetry along the $z$ axis, the integration in $\lambda'$ is
straightforward:

@@

\begin{equation}
    g_{xx}(r) = g_{yy}(r) = -\frac{V_\text{sh}(r)}{r^2},
\end{equation}

@@

From Eq. \ref{eq:shell-pot} we can obtain expressions of the gravitational potential for
different density functions.
For a linear density function
